### 2016-03-06 text/Chapter04.tex
Move stuff in chapter 4, improve introductions

Successful programming language abstractions need to generalize a wide range of recurring
problems while capturing the key commonalities. These two aims are typically in opposition -- 
more general abstractions are less powerful, while less general abstractions cannot be
used as often.

In the previous chapter, we outlined a number of systems that capture how computations
access the environment in which they are executed. We identified two kinds of systems --
\emph{flat systems} capturing whole-context properties and \emph{structural systems} capturing 
per-variable properties. As we show in Section~X, the systems can be further unified using a 
single abstraction, but such abstraction is  \emph{less powerful} -- \ie~its generality hides 
useful properties that we can see when we consider the systems separately. For this reason, we 
discuss \emph{flat coeffects} and \emph{structural coeffects} separately.

In this and the next chapter, we discuss the type system and the semantics of flat coeffect 
systems, respectively. In this chapter, we develop a parameterized type system for flat coeffect
systems and we study its general syntactic properties. We also consider variations of the type system
that resolve the ambiguity of coeffectful lambda abstraction for concrete instances of the system.
In the next chapter, we give operational meaning of concrete coeffect languages using the unified
system and we discuss their safety.

@@

\item The type system permits multiple typing derivation for certain programs due to the ambiguity
  inherent in conextual lambda abstraction rule. In Section~\ref{sec:flat-unique}, we discuss
  variations of the type system that resolve the ambiguity and give a unique typing derivation
  for the three coeffect systems covered in this chapter.

\item We discuss syntactic properties of the calculus, covering type-pre\-ser\-vation for call-by-name
  and call-by-value reduction (Section~\ref{sec:flat-syntax}). We also extend the calculus
  with subtyping and pairs (Section~\ref{sec:flat-exts}). These two sections motivate the laws
  of the flat coeffect algebra.
\end{itemize}

@@

In the previous chapter, we looked at three important examples of systems that track whole-context 
properties. The type systems for whole-context liveness (Section~\ref{sec:applications-flat-live}) 
and whole-context dataflow (Section~\ref{sec:applications-flat-dataflow}) have a very similar 
structure. First, their lambda abstraction duplicates the requirements. Given a body with context 
requirements $\cclrd{r}$, the declaration site context \emph{as well as} the function arrow are
annotated with $\cclrd{r}$. Second, their application arises from the combination of \emph{sequential} 
and \emph{pointwise} composition.

The system for tracking of implicit parameters and similar (Section~\ref{sec:applications-flat-impl})
differ in two ways. In lambda abstraction, they split the context requirements between the 
declaration site and the call site and they use only a single operator on the indices, typically $\cup$.

Despite the differences, the systems fit the same unified framework. This becomes apparent when
we consider the categorical structure (Section~\ref{sec:flat-semantics}). However, rather than 
starting from the semantics, we first explain how the systems can be unified syntactically
(Section~\ref{sec:flat-calculus-lambda}) and then provide the semantics as a justification.

The development in this chapter can be seen as a counterpart to the well-known development of 
\emph{effect systems} \cite{effects-gifford}. The Chapter~\ref{ch:semantics} then links \emph{coeffects} 
with \emph{comonads} in the same way in which effect systems can be linked with monads
\cite{monad-notions}. The syntax and type system of the flat coeffect calculus follows 
a similar style as effect systems \cite{effects-polymorphic,effects-talpin-et-al}, but differs
in the structure, as explained in the previous chapter, most importantly in lambda abstraction
(the relationship with monads is further discussed in Section~\ref{sec:flat-related}).

The flat coeffect calculus is defined in terms of \emph{flat coeffect algebra}, which defines
the structure of context annotations, such as $\cclrd{r}, \cclrd{s}, \cclrd{t}$. These can be
sets of implicit parameters, versions represented as integers or other values. The expressions of 
the calculus are those of the $\lambda$-calculus with \emph{let} binding. We also include a type 
\ident{num} as an example of a concrete base type with numerical constants written as $n$:
%
\begin{equation*}
\begin{array}{rcl}
e &::=& x \sep n \sep \lambda x:\tau.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
\tau &::=& \ident{num} \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
\end{array}
\end{equation*}
%
Note that the lambda abstrction in the syntax is written in the Church-style and requires a type
annotation. This will be used in Section~\ref{sec:flat-unique} where we discuss how to find a 
unique typing derivation for context-aware computations. Using Church-style lambda abstraction,
we can directy focus on the more interesting problem of finding unique \emph{coeffect annotations} 
rather than solving the problem of type reconstruction.

@@

@@

Before looking at the type system in Figure~\ref{fig:flat-types}, let us clarify how the rules
should be understood. The coeffect calculus provides both analysis of context dependence (type 
system) and semantics for context (how it is propagated). These two aspects provide different
ways of reading the judgements $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ and the typing rules
used to define it.

\begin{itemize}
\item \textsc{Analysis of context dependence.}
Syntactically, coeffect annotations $\cclrd{r}$ model \emph{context requirements}. This means
we can over-approximate them and require more than is actually needed at runtime.

Syntactically, the typing rules should be read top-down. In function application, the context 
requirements of multiple assumptions (arising from two sub-expressions) are \emph{merged}; in 
lambda abstraction, the requirements of a single expression (the body) are split between
the declaration site and the call site.

@@

The reason for this asymmetry follows from the fact that the context appears in a \emph{negative
position} in the semantic model (Section~\ref{sec:flat-semantics}). It means that we need to be
careful about using the words \emph{split} and \emph{merge}, because they can be read as meaning
exactly the opposite things. To disambiguate, we always use the term \emph{context requirements} 
when using the syntactic view, especially in the rest of Chapter~\ref{chf:flat}, and 
\emph{context capabilities} or just \emph{available context} when using the semantic view, 
especially in Chapter~\ref{ch:semantics}.

@@

To make the flat coeffect system general enough, the algebra consists of three operations.
Two of them, $\cseq$ and $\cpar$, represent the \emph{sequential} and \emph{pointwise} composition, 
which are mainly used in function application. The third operator, $\czip$ is used in lambda 
abstraction and represents \emph{splitting} of context requirements (or, semantically, \emph{merging} 
of available context capabilities).

@@

\item For the $\czip$ operation, we require associativity and idempotence. The idempotence
requirement makes it possible to duplicate the coeffects and place the same requirement on both
call site and declaration site. Using the example from Section~\ref{sec:flat-calculus-lambda},
this guarantees that the rule (\emph{df1}) is not a special case, but can always be derived 
from (\emph{df2}). In some cases, the operator forms a monoid with the unit being the greatest 
element of the set $\C$. 
\end{itemize}

@@

Using the syntactic reading, the two operators represent \emph{merging} and \emph{splitting} of context 
requirements -- in the (\emph{abs}) rule, $\czip$ appears in the assumption and the combined context 
requirements of the body are split between two positions in the conclusions; in the (\emph{app}) rule, 
$\cpar$ appears in the conclusion and combines two context requirements from the assumptions.

@@

@@

\subsection{Flat coeffect types}
\label{sec:flat-calculus-types}

@@

The (\emph{abs}) rule is defined as discussed in Section~\ref{sec:flat-calculus-lambda}. The
body is annotated with context requirements $\cclrd{r} \,\czip\, \cclrd{s}$, which are then split
between the context-requirements on the declaration site $\cclrd{r}$ and context-requirements on
the call site $\cclrd{s}$. Examples of the $\czip$ operator are discussed in the next section.

In function application (\emph{app}), context requirements of both expressions and the 
function are combined. As discussed in Chapter~\ref{ch:applications}, the pointwise composition
$\cpar$ is used to combine the context requirements of the expression representing a function 
$\cclrd{r}$ and the context requirements of the argument, sequentially composed with the 
context-requirements of the function $\cclrd{s}\, \cseq \,\cclrd{t}$.

@@

As discussed in Chapter~\ref{sec:applications}, the lambda abstraction rule for coeffect systems
differs from the rule for effect systems in that it does not delay all context requirements. 
In case of implicit parameters (Section~\ref{sec:applications-flat-impl}), the requirements
can be satisfied either by the call-site or by the declaration-site. In case of dataflow and
liveness, the rule discussed in Section~\ref{sec:flat-calculus} reintroduces similar ambiguity
because it allows multiple valid typing derivations.

@@

For implicit parameters we follow the behaviour implemented by Haskell \cite{app-implicit-parameters}
where function abstraction captures all parameters that are available at the declaration-site and 
places all other requirements on the call-site. For the example in the introduction, this means
that the body of $f$ captures the value of \ident{?p} available from the declaration-site
and $f$ will be typed as a function requiring no parameters (coeffect $\cclrd{\emptyset}$). The 
program thus evaluates to $1$.

@@

Briefly, the implementation collects constraints on the coeffects and then finds the smallest sets 
of implicit parameters to satisfy the constraints. We still need to track implicit parameters in 
scope $\cclrd{\Delta}$, but the rest of the (\emph{abs}) rule from the implementation is close
to the one from Figure~\ref{fig:flat-types}:
%
\begin{equation*}
\tyrule{abs}
  {\coctx{\Gamma, x\!:\!\tau_1;\cclrd{\Delta}}{\cclrd{t}} \vdash e : \tau_2 ~|~ C}
  {\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{r}} \vdash \lambda x\!:\!\tau_1.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 ~|~
    C\cup\{\cclrd{t}=\cclrd{r} \,\czip\, \cclrd{s}, \cclrd{r}=\cclrd{\Delta} \}} 
\end{equation*}
%
Given a typing derivation for the body that produced constraints $C$, we generate an additional
constraint that restricts $\cclrd{r}$ (declaration-site requirements) to those available in the
current static scope $\cclrd{\Delta}$. The constraint satisfaction algorithm then finds 
the minimal set $\cclrd{s}$ which is $\cclrd{t}\cclrd{\setminus \Delta}$.

@@

The notion of \emph{value} in coeffect systems differs from the usual syntactic understanding.
A function $(\lambda x.e)$ does not defer all context requirements of the body $e$ and may have
immediate context requirements. Thus we say that $e$ is a value if it is a value in the usual
sense \emph{and} has not immediate context requirements. We define this formally in 
Section~\ref{sec:flat-syntax-cbv}.

The call-by-value evaluation strategy preserves typing for a wide range of flat coeffect calculi, 
including all our three examples. However, it is rather weak -- in order to use it, the concrete 
semantics needs to provide a way for reducing context-dependent term $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
to a value, \ie~a term $\coctx{\Gamma}{\cunit} \vdash e' : \tau$ with no context requirements.

@@

As discussed in the previous section, call-by-value reduction can be used for most flat coeffect
calculi, but it provides a very weak general model \ie~the hard work of reducing context-dependent
term to a \emph{value} has to be provided for each system. Syntactic values are defined
in the usual way:
%
\begin{equation*}
\begin{array}{llll}
v\in\mathit{SynVal}   & v &::=& x \sep c \sep (\lambda x.e)\\
n\in\mathit{NonVal}   & n &::=& e_1\;e_2 \sep \kvd{let}~x=e_1~\kvd{in}~e_2\\
e\in\mathit{Expr}     & e &::=& v \sep n 
\end{array}
\end{equation*}
%
The syntactic form \emph{SynVal} captures syntactic values, but a context-dependency-free value in 
coeffect calculus cannot be defined purely syntactically, because a function $(\lambda x.e)$ does
not automatically defer all context requirements.

@@

\noindent
The call-by-value substitution substitutes a value, with context requirements $\cunit$, for a 
variable, whose access is also annotated with $\cunit$. Thus, it does not affect the type and
context requirements of the term:

\begin{lemma}[Call-by-value substitution]
\label{thm:flat-subst-cbv}
In a flat coeffect calculus with a coeffect algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$,
given a value $\coctx{\Gamma}{\cunit} \vdash v : \sigma$ and an expression
$\coctx{\Gamma,  x\!:\!\sigma}{ \cclrd{r}  } \vdash e : \tau$, then substituting $v$ for $x$ does not
change the type and context requirements, that is $\coctx{\Gamma}{ \cclrd{r} } \vdash \subst{e}{x}{v} : \tau$.
\end{lemma}
\begin{proof}
By induction over the type derivation, using the fact that $x$ and $v$ are annotated
with $\cunit$ and that $\Gamma$ is treated as a set in the flat calculus.
\end{proof}

\noindent
The substitution lemma~\ref{thm:flat-subst-cbv} holds for all flat coeffect systems. However, 
proving that call-by-value reduction preserves typing requires an additional constraint on the
flat coeffect algebra, which relates the $\czip$ and $\cpar$ operations. This is captured
by the (\emph{approximation}) property:
%
\begin{equation*}
\cclrd{r}\,\czip\,\cclrd{t}~\;\cleq~\;\cclrd{r}\,\cpar\,\cclrd{t}
\hspace{10em}(\textit{approximation})
\end{equation*}
%
Intuitively, this specifies that the $\czip$ operation (splitting of context requirements) 
under-approximates the actual context capabilities while the $\cpar$ operation (combining of
context requirements) over-approximates the actual context requirements.

@@

Semantically, the coeffect over-approximates the actual requirements -- at run-time, the code does not 
actually access a previous value of the argument $x$. This cannot be captured by a flat coeffect 
system, but it can be captured using the structural system discussed in Chapter~\ref{ch:structural}.

@@

\noindent
Liveness is an example of top-pointed coeffects as variables are annotated with 
$\ident{L}$ and $\ident{D} \leq \ident{L}$, while implicit parameters and dataflow are examples
of bottom-pointed coeffects. For top-pointed flat coeffects, the substitution lemma holds without 
additional requirements:

@@

In a bottom pointed coeffect system, substituting $e$ for $x$ increases the context 
requirements. However, if the system satisfies the strong condition that $\czip = \cseq = \cpar$ 
then the context requirements arising from the substitution can be associated with the context
$\Gamma$, leaving the context requirements of a function value unchanged. As a result, substitution 
does not break soundness as in effect systems. The requirement $\czip = \cseq = \cpar$ holds for 
our implicit parameters example (all three operators are a set union) and for other set-like 
coeffects. It allows the following substitution lemma:

@@

\noindent
The flat system discussed here is \emph{flexible enough} to let us always re-associate new context 
requirements (arising from the substitution) with the free-vari\-able context. In contrast, the 
structural system discussed in Chapter~\ref{ch:structural} is \emph{precise enough} to keep the 
coeffects associated with individual variables, thus preserving typing in a complementary way.

@@

Recall that $\eta$-expansion turns $e$ into $\lambda x.e~x$. In the case of liveness, the 
expression $e$ may require no variables (both immediate and latent coeffects are marked as
$\ident{D}$). However, the resulting expression $\lambda x.e~x$ accesses a variable, 
marking the context and function argument as live. In case of dataflow, the immediate coeffects 
are made larger by the lambda abstraction -- the context requirements of the function value are 
imposed on the declaration site of the new lambda abstraction. We remedy this limitation in 
the next chapter.

@@

\begin{theorem}[Type preservation of $\eta$-expansion]
In a bottom-pointed flat coeffect calculus with an algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ 
where $\czip = \cpar$, if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2$ 
and $e \rightarrow_\eta e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 $.
\end{theorem}
\begin{proof}
The following derivation shows that $\lambda x.f~x$ has the same type as $f$:
\begin{equation*}
\inference
  { \inference
    { \coctx{\Gamma}{\cclrd{r}} \vdash f : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 &
      \coctx{x\!:\!\tau_1}{\cunit} \vdash x\!:\!\tau_1 }
    { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r} \;\cpar\; (\cunit \,\cseq\, \cclrd{s}) } \vdash f~x : \tau_2 } }
  { \inference
    { \inference
      { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r} \;\cpar\; \cclrd{s} } \vdash f~x : \tau_2 }
      { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r} \;\czip\; \cclrd{s} } \vdash f~x : \tau_2 } }
    { \coctx{\Gamma}{ \cclrd{r} } \vdash \lambda x.f~x : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 } }
\end{equation*}
%
The derivation starts with the expression $e$ and derives the type for $\lambda x.e~x$. The
application yields context requirements $\cclrd{r}\,\cpar\,\cclrd{s}$. In order to recover the
original typing, this must be equal to $\cclrd{r}\,\czip\,\cclrd{s}$. Note that the derivation
is showing just one possible typing -- the expression $\lambda x.e~x$ has other types -- but
this is sufficient for showing type preservation.
\end{proof}

@@

In Section~\ref{sec:flat-calculus-lambda}, we discussed how to reconcile two typings for
lambda abstraction -- for implicit parameters, the lambda function needs to split context 
requirements using $\cclrd{r} \cup \cclrd{s}$, but for dataflow and liveness it suffices to 
duplicate the requirement $\cclrd{r}$ of the body. We introduced the $\czip$ operation as 
a way of providing the additional abstraction.

@@

\paragraph{Simplified abstraction.}
Recall that $(\C, \czip)$ is a band, that is, $\czip$  is idempotent and associative. The 
idempotence means that the context requirements of the body can be required from both the 
declaration site and the call site. Thus, the (\emph{idabs}) typing is valid. For reference,
we repeat the earlier definitions here:

>>>>

In the previous chapter, we outlined a number of systems that capture how computations
access the environment in which they are executed. We identified two kinds of systems --
\emph{flat systems} capturing whole-context properties and \emph{structural systems} capturing 
per-variable properties. As we show in Section~X, the systems can be further unified using a 
single abstraction, but such abstraction is \emph{less powerful} -- \ie~its generality hides 
useful properties that we can see when we consider the systems separately. For this reason, we 
discuss \emph{flat coeffects} (Chapter~\ref{ch:flat} and Chapter~\ref{ch:semantics}) and 
\emph{structural coeffects} (Chapter~\ref{ch:structural}) separately.

In this chapter, we develop a \emph{flat coeffect calculus} that provides a type system for 
tracking per-context properties of context-aware programming languages. The \emph{coeffect 
calculus} captures the shared properties of such languages. It is parameterized by a \emph{flat
coeffect algebra} and can be instantiated to track implicit parameters, liveness and number of
required past values in dataflow languages.

To capture contextual properties in full generality, the flat coeffect calculus permits multiple 
valid typing derivations for a given term. To resolve the ambiguity arising from such generality, 
each concrete context-aware language is also equipped with an algorithm for choosing a unique 
typing derivation. In the next chapter, we give operational meaning for concrete coeffect 
languages based on the flat coeffect calculus and we discuss their safety.

@@

\item The coeffect calculus permits multiple typing derivations due to the ambiguity inherent 
  in the conextual lambda abstraction rule. Each concrete context-aware language based on the 
  coeffect calculus may resolve this ambiguity differently. We show the procedure for choosing
  unique typing derivation for our three examples (Section~\ref{sec:flat-unique}).

\item We discuss syntactic properties of the calculus, covering type-pre\-ser\-vation for call-by-name
  and call-by-value reduction (Section~\ref{sec:flat-syntax}). We also extend the calculus
  with subtyping and pairs (Section~\ref{sec:flat-exts}). These two sections justify the laws
  of the flat coeffect algebra.
\end{itemize}

@@

In the previous chapter, we looked at three examples of systems that track whole-context 
properties. The type systems for whole-context liveness (Section~\ref{sec:applications-flat-live}) 
and whole-context dataflow (Section~\ref{sec:applications-flat-dataflow}) have a similar 
structure in two ways. First, their lambda abstraction duplicates the demands. Given a body with 
context demands $\cclrd{r}$, the declaration site context \emph{as well as} the function arrow are
annotated with $\cclrd{r}$. Second, the context demands in the type systems are combined using 
two different operators (representing sequential and pointwise operations).

The system for tracking of implicit parameters (Section~\ref{sec:applications-flat-impl})
differs. In lambda abstraction, it splits the context demands between the 
declaration site and the call site and they use only a single operator on the indices, typically $\cup$.

Despite the differences, the systems fit the same framework. This becomes apparent when
we consider the categorical structure (Section~\ref{sec:flat-semantics}). However, rather than 
starting from the semantics, we first explain how the systems can be unified syntactically
(Section~\ref{sec:flat-calculus-lambda}) and then provide the semantics as a justification.

The development in this chapter can be seen as a counterpart to the well-known development of 
\emph{effect systems} \cite{effects-gifford}. The Chapter~\ref{ch:semantics} then links \emph{coeffects} 
with \emph{comonads} in the same way in which effect systems can be linked with monads
\cite{monad-notions}. The syntax and type system of the flat coeffect calculus follows 
a similar style as effect systems \cite{effects-polymorphic,effects-talpin-et-al}, but differs
in the structure of lambda abstraction as discussed in Section~\ref{sec:applications-structure-lam}
(the relationship with monads is further discussed in Section~\ref{sec:semantics-related}).

@@

The \emph{flat coeffect calculus} is parameterized by a \emph{flat coeffect algebra}, which defines
the structure of context annotations, written as $\cclrd{r}, \cclrd{s}, \cclrd{t}$. These can be
sets of implicit parameters, versions represented as integers or other values.

Typing judgements the attach coeffect annotations $\cclrd{r}, \cclrd{s}, \cclrd{t}$ to the context
and are written as $\coctx{\Gamma}{\cclrd{r}}\vdash e:\tau$.

@@

Before looking at the typing rules, let us clarify how the typing judgements
should be understood. The coeffect calculus provides both analysis of context dependence (type 
system) and semantics for context (how it is propagated). These two aspects provide different
ways of reading the judgements $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ and the typing rules
used to define it.

\begin{itemize}
\item \textsc{Analysis of context dependence.}
Syntactically, coeffect annotations $\cclrd{r}$ model \emph{context demands}. This means
we can over-approximate them and require more than is actually needed at runtime.

Syntactically, the typing rules are best read top-down. In function application, the context 
demands of multiple assumptions (arising from two sub-expressions) are \emph{merged}; in 
lambda abstraction, the demands of a single expression (the body) are split between
the declaration site and the call site.

@@

The reason for this asymmetry follows from the fact that the context appears in a \emph{negative
position} in the semantic model (Section~\ref{sec:semantics-theory}). It means that we need to be
careful about using the words \emph{split} and \emph{merge}, because they can be read as meaning
exactly the opposite things. To disambiguate, we always use the term \emph{context demands} 
when using the syntactic view, especially in the rest of Chapter~\ref{chf:flat}, and 
\emph{context capabilities} or just \emph{available context} when using the semantic view, 
especially in Chapter~\ref{ch:semantics}.

@@

To make the flat coeffect system general enough, the algebra consists of three operations.
Two of them, $\cseq$ and $\cpar$, represent the \emph{sequential} and \emph{pointwise} composition, 
which are mainly used in function application. The third operator, $\czip$ is used in lambda 
abstraction and represents \emph{splitting} of context demands (or, semantically, \emph{merging} 
of available context capabilities).

@@

\item For the $\czip$ operation, we require associativity and idempotence. The idempotence
demand makes it possible to duplicate the coeffects and place the same demand on both
call site and declaration site. Using the example from Section~\ref{sec:flat-calculus-lambda},
this guarantees that the rule (\emph{df1}) is not a special case, but can always be derived 
from (\emph{df2}). In some cases, the operator forms a monoid with the unit being the greatest 
element of the set $\C$. 
\end{itemize}

@@

Using the syntactic reading, the two operators represent \emph{merging} and \emph{splitting} of context 
demands -- in the (\emph{abs}) rule, $\czip$ appears in the assumption and the combined context 
demands of the body are split between two positions in the conclusions; in the (\emph{app}) rule, 
$\cpar$ appears in the conclusion and combines two context demands from the assumptions.

@@

\subsection{Language syntax}
\label{sec:flat-calculus-syntax}

The expressions of 
the calculus are those of the $\lambda$-calculus with \emph{let} binding. We also include a type 
\ident{num} as an example of a concrete base type with numerical constants written as $n$:
%
\begin{equation*}
\begin{array}{rcl}
e &::=& x \sep n \sep \lambda x:\tau.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
\tau &::=& \ident{num} \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
\end{array}
\end{equation*}
%
Note that the lambda abstrction in the syntax is written in the Church-style and requires a type
annotation. This will be used in Section~\ref{sec:flat-unique} where we discuss how to find a 
unique typing derivation for context-aware computations. Using Church-style lambda abstraction,
we can directy focus on the more interesting problem of finding unique \emph{coeffect annotations} 
rather than solving the problem of type reconstruction.

@@

\subsection{Type system}
\label{sec:flat-calculus-types}

@@

The (\emph{abs}) rule is defined as discussed in Section~\ref{sec:flat-calculus-lambda}. The
body is annotated with context demands $\cclrd{r} \,\czip\, \cclrd{s}$, which are then split
between the context-demands on the declaration site $\cclrd{r}$ and context-demands on
the call site $\cclrd{s}$. Examples of the $\czip$ operator are discussed in the next section.

In function application (\emph{app}), context demands of both expressions and the 
function are combined. As discussed in Chapter~\ref{ch:applications}, the pointwise composition
$\cpar$ is used to combine the context demands of the expression representing a function 
$\cclrd{r}$ and the context demands of the argument, sequentially composed with the 
context-demands of the function $\cclrd{s}\, \cseq \,\cclrd{t}$.

@@

As discussed in Chapter~\ref{sec:applications}, the lambda abstraction rule for coeffect systems
differs from the rule for effect systems in that it does not delay all context demands. 
In case of implicit parameters (Section~\ref{sec:applications-flat-impl}), the demands
can be satisfied either by the call-site or by the declaration-site. In case of dataflow and
liveness, the rule discussed in Section~\ref{sec:flat-calculus} reintroduces similar ambiguity
because it allows multiple valid typing derivations.

@@

For implicit parameters we follow the behaviour implemented by Haskell \cite{app-implicit-parameters}
where function abstraction captures all parameters that are available at the declaration-site and 
places all other demands on the call-site. For the example in the introduction, this means
that the body of $f$ captures the value of \ident{?p} available from the declaration-site
and $f$ will be typed as a function requiring no parameters (coeffect $\cclrd{\emptyset}$). The 
program thus evaluates to $1$.

@@

Briefly, the implementation collects constraints on the coeffects and then finds the smallest sets 
of implicit parameters to satisfy the constraints. We still need to track implicit parameters in 
scope $\cclrd{\Delta}$, but the rest of the (\emph{abs}) rule from the implementation is close
to the one from Figure~\ref{fig:flat-types}:
%
\begin{equation*}
\tyrule{abs}
  {\coctx{\Gamma, x\!:\!\tau_1;\cclrd{\Delta}}{\cclrd{t}} \vdash e : \tau_2 ~|~ C}
  {\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{r}} \vdash \lambda x\!:\!\tau_1.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 ~|~
    C\cup\{\cclrd{t}=\cclrd{r} \,\czip\, \cclrd{s}, \cclrd{r}=\cclrd{\Delta} \}} 
\end{equation*}
%
Given a typing derivation for the body that produced constraints $C$, we generate an additional
constraint that restricts $\cclrd{r}$ (declaration-site demands) to those available in the
current static scope $\cclrd{\Delta}$. The constraint satisfaction algorithm then finds 
the minimal set $\cclrd{s}$ which is $\cclrd{t}\cclrd{\setminus \Delta}$.

@@

The notion of \emph{value} in coeffect systems differs from the usual syntactic understanding.
A function $(\lambda x.e)$ does not defer all context demands of the body $e$ and may have
immediate context demands. Thus we say that $e$ is a value if it is a value in the usual
sense \emph{and} has not immediate context demands. We define this formally in 
Section~\ref{sec:flat-syntax-cbv}.

The call-by-value evaluation strategy preserves typing for a wide range of flat coeffect calculi, 
including all our three examples. However, it is rather weak -- in order to use it, the concrete 
semantics needs to provide a way for reducing context-dependent term $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
to a value, \ie~a term $\coctx{\Gamma}{\cunit} \vdash e' : \tau$ with no context demands.

@@

As discussed in the previous section, call-by-value reduction can be used for most flat coeffect
calculi, but it provides a very weak general model \ie~the hard work of reducing context-dependent
term to a \emph{value} has to be provided for each system. Syntactic values are defined
in the usual way:
%
\begin{equation*}
\begin{array}{llll}
v\in\mathit{SynVal}   & v &::=& x \sep c \sep (\lambda x.e)\\
n\in\mathit{NonVal}   & n &::=& e_1\;e_2 \sep \kvd{let}~x=e_1~\kvd{in}~e_2\\
e\in\mathit{Expr}     & e &::=& v \sep n 
\end{array}
\end{equation*}
%
The syntactic form \emph{SynVal} captures syntactic values, but a context-dependency-free value in 
coeffect calculus cannot be defined purely syntactically, because a function $(\lambda x.e)$ does
not automatically defer all context demands.

@@

\noindent
The call-by-value substitution substitutes a value, with context demands $\cunit$, for a 
variable, whose access is also annotated with $\cunit$. Thus, it does not affect the type and
context demands of the term:

\begin{lemma}[Call-by-value substitution]
\label{thm:flat-subst-cbv}
In a flat coeffect calculus with a coeffect algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$,
given a value $\coctx{\Gamma}{\cunit} \vdash v : \sigma$ and an expression
$\coctx{\Gamma,  x\!:\!\sigma}{ \cclrd{r}  } \vdash e : \tau$, then substituting $v$ for $x$ does not
change the type and context demands, that is $\coctx{\Gamma}{ \cclrd{r} } \vdash \subst{e}{x}{v} : \tau$.
\end{lemma}
\begin{proof}
By induction over the type derivation, using the fact that $x$ and $v$ are annotated
with $\cunit$ and that $\Gamma$ is treated as a set in the flat calculus.
\end{proof}

\noindent
The substitution lemma~\ref{thm:flat-subst-cbv} holds for all flat coeffect systems. However, 
proving that call-by-value reduction preserves typing requires an additional constraint on the
flat coeffect algebra, which relates the $\czip$ and $\cpar$ operations. This is captured
by the (\emph{approximation}) property:
%
\begin{equation*}
\cclrd{r}\,\czip\,\cclrd{t}~\;\cleq~\;\cclrd{r}\,\cpar\,\cclrd{t}
\hspace{10em}(\textit{approximation})
\end{equation*}
%
Intuitively, this specifies that the $\czip$ operation (splitting of context demands) 
under-approximates the actual context capabilities while the $\cpar$ operation (combining of
context demands) over-approximates the actual context demands.

@@

Semantically, the coeffect over-approximates the actual demands -- at run-time, the code does not 
actually access a previous value of the argument $x$. This cannot be captured by a flat coeffect 
system, but it can be captured using the structural system discussed in Chapter~\ref{ch:structural}.

@@

\noindent
Liveness is an example of top-pointed coeffects as variables are annotated with 
$\ident{L}$ and $\ident{D} \leq \ident{L}$, while implicit parameters and dataflow are examples
of bottom-pointed coeffects. For top-pointed flat coeffects, the substitution lemma holds without 
additional demands:

@@

In a bottom pointed coeffect system, substituting $e$ for $x$ increases the context 
demands. However, if the system satisfies the strong condition that $\czip = \cseq = \cpar$ 
then the context demands arising from the substitution can be associated with the context
$\Gamma$, leaving the context demands of a function value unchanged. As a result, substitution 
does not break soundness as in effect systems. The demand $\czip = \cseq = \cpar$ holds for 
our implicit parameters example (all three operators are a set union) and for other set-like 
coeffects. It allows the following substitution lemma:

@@

\noindent
The flat system discussed here is \emph{flexible enough} to let us always re-associate new context 
demands (arising from the substitution) with the free-vari\-able context. In contrast, the 
structural system discussed in Chapter~\ref{ch:structural} is \emph{precise enough} to keep the 
coeffects associated with individual variables, thus preserving typing in a complementary way.

@@

Recall that $\eta$-expansion turns $e$ into $\lambda x.e~x$. In the case of liveness, the 
expression $e$ may require no variables (both immediate and latent coeffects are marked as
$\ident{D}$). However, the resulting expression $\lambda x.e~x$ accesses a variable, 
marking the context and function argument as live. In case of dataflow, the immediate coeffects 
are made larger by the lambda abstraction -- the context demands of the function value are 
imposed on the declaration site of the new lambda abstraction. We remedy this limitation in 
the next chapter.

@@

\begin{theorem}[Type preservation of $\eta$-expansion]
In a bottom-pointed flat coeffect calculus with an algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ 
where $\czip = \cpar$, if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2$ 
and $e \rightarrow_\eta e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 $.
\end{theorem}
\begin{proof}
The following derivation shows that $\lambda x.f~x$ has the same type as $f$:
\begin{equation*}
\inference
  { \inference
    { \coctx{\Gamma}{\cclrd{r}} \vdash f : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 &
      \coctx{x\!:\!\tau_1}{\cunit} \vdash x\!:\!\tau_1 }
    { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r} \;\cpar\; (\cunit \,\cseq\, \cclrd{s}) } \vdash f~x : \tau_2 } }
  { \inference
    { \inference
      { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r} \;\cpar\; \cclrd{s} } \vdash f~x : \tau_2 }
      { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r} \;\czip\; \cclrd{s} } \vdash f~x : \tau_2 } }
    { \coctx{\Gamma}{ \cclrd{r} } \vdash \lambda x.f~x : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 } }
\end{equation*}
%
The derivation starts with the expression $e$ and derives the type for $\lambda x.e~x$. The
application yields context demands $\cclrd{r}\,\cpar\,\cclrd{s}$. In order to recover the
original typing, this must be equal to $\cclrd{r}\,\czip\,\cclrd{s}$. Note that the derivation
is showing just one possible typing -- the expression $\lambda x.e~x$ has other types -- but
this is sufficient for showing type preservation.
\end{proof}

@@

In Section~\ref{sec:flat-calculus-lambda}, we discussed how to reconcile two typings for
lambda abstraction -- for implicit parameters, the lambda function needs to split context 
demands using $\cclrd{r} \cup \cclrd{s}$, but for dataflow and liveness it suffices to 
duplicate the demand $\cclrd{r}$ of the body. We introduced the $\czip$ operation as 
a way of providing the additional abstraction.

@@

\paragraph{Simplified abstraction.}
Recall that $(\C, \czip)$ is a band, that is, $\czip$  is idempotent and associative. The 
idempotence means that the context demands of the body can be required from both the 
declaration site and the call site. Thus, the (\emph{idabs}) typing is valid. For reference,
we repeat the earlier definitions here:
